Fractional Decomposition Tree Algorithm: A tool for studying the integrality gap of Integer Programs

### 2020-06-12 FDT2020/Introduction.tex
Adding required Sandia line (necessary to pass internal review), making a section a subsection, etc
Work on introduction

In combinatorial optimization the aim is to find the optimal solution in a discrete and
usually finite yet large set of solutions. For many specific combinatorial optimization problems such a solution can be found efficiently. For many others, finding optimal or in many cases near optimal solutions is NP-hard. A common approach to deal with such problems is relaxing the discrete solution set into a continuous set, where the optimization problem becomes tractable. Obtaining feasible solutions by means of such a relaxation requires an additional step of rounding the potentially fractional solution of the continuous relaxation into integer solutions.
}\fi
\cindy{TODO: remove ``recall that'' for anything we haven't already defined in this paper.  Even though it's likely anyone reading this will care about IPs and integrality gaps, we should not assume they already know things we define the way we define them.}

\section{Introduction}
In this paper we focus on finding solutions to general Integer Linear Programs (IP). Integer Programming (and more generally Mixed Integer Linear Programming) can be used to model many practical optimization problems including scheduling, logistics and resource allocation.

A common approach to Integer programming is considering their linear relaxation. For problems such as the \textsc{Minimum Cost Spanning Tree Problem} there are linear programming relaxations whose basic feasible solutions coincide with  integral solutions, i.e. spanning trees. For other problems the value of the linear programming relaxation provides a bound (lower bound for a minimization problem and upper bound for a maximization problem) on the optimal solution. A common and successful approach is to round these (potentially) fractional solutions into integer solutions for the optimization problem at hand. The Integrality gap of a linear relaxation of an integer programming problem is the worst case ratio between the objective values of the discrete problem and the continuous problem. Equivalently, the integrality gap of the linear programming relaxation is a limit to the rounding approach: rounding a fractional solution into an integer solution incurs
a multiplicative cost proportional to the integrality gap. In this dissertation we study integrality gaps for different combinatorial optimization problems and introduce new rounding algorithms that imply bounds on their respective integrality gaps.

@@

A polynomial time algorithm for proving an upper bound on integrality gap is called an LP-based approximation algorithm. For many well studied problems, we still do not know the exact integrality gap and the gap between the best known lower bound and the upper bound on the integrality gap are open. In some cases, there are known upper bounds, yet there is no known approximation algorithm, meaning that the proofs do not yield polynomial time algorithms.
}\fi 
The set of feasible points for a pure IP (henceforth IP) is the set
\begin{equation}
S(A,b)= \{x\in \mathbb{Z}^{n}\;:\; Ax\geq b\}  \label{S}.
\end{equation}
If we drop the integrality constraints, we have the linear relaxation of set $S(A,b)$,
\begin{equation}
P(A,b) = \{x\in \mathbb{R}^{n}\;:\; Ax\geq b\}. \label{P}
\end{equation}
Let $I=(A,b)$ denote an instance. Then $S(I)$ and $P(I)$ denote $S(A,b)$ and $P(A,b)$, respectively. Given a linear objective function $c$, an IP is $\min \;\{cx:\; x \in S(I)\}$. It is  NP-hard even to determine if an IP instance has a feasible solution~\cite{GareyJohnson}. However, intelligent branch-and-bound strategies allow commercial and open-source MILP solvers to give exact solutions (or near-optimal solutions with a provable bound) to many specific instances of NP-hard combinatorial optimization problems.

Relaxing the integrality constraints gives the polynomial-time-solvable linear programming relaxation: $\min \;\{cx:\;x\in P(I) \}$.  The optimal value of this linear program (LP), denoted $z_{\lp}(I,c)$, is a lower bound on the optimal value for the IP, denoted $z_{\IP}(I,c)$. The solutions can also provide some useful global structure, even though the fractional values might not be directly meaningful.

Many researchers (see \cite{davids,vazirani}) have developed polynomial time LP-based approximation algorithms that find solutions for special classes of IPs whose cost are provably smaller than $C\cdot z_{LP}(I,c)$. The approximation factor $C$ can be a constant or depend on the input parameters of the IP, e.g. $O(\log(n))$ where $n$ is the number of variables in the formulation of the IP (the dimension of the problem). However, for many combinatorial optimization problems there is a limit to such techniques based on LP relaxations, represented  by the {integrality gap} of the IP formulation. The integrality gap $g(I)$ for instance $I$ is defined to be $g(I)= \max_{c\geq 0}\frac{z_{IP}(I,c)}{z_{LP}(I,c)}$. \cindy{Does the integrality gap always require positive objective coefficients?  Or is that an extra restriction in our definition?  Is there a reference that has that restriction?} For example consider the minimum cost 2-edge-connected multi-subgraph problem (2EC): Given a graph $G=(V,E)$ and $c\in \mathbb{R}^E_{\geq 0}$, 2EC asks for the minimum cost 2-edge-connected multi-subgraph of $G$. A linear programming relaxation for this problem known as the subtour elimination relaxation  is
 \begin{equation}\min \{cx: \sum_{e\in \delta(U)}x_e \geq 2 \mbox{ for } \emptyset \subsetneq U \subsetneq V,\; x\in [0,2]^{E}\}. 
 \end{equation}
 In this case the instance-specific integrality gap is the integrality gap of the subtour-elimination relaxation for the 2EC on graph with $n$ vertices. \cindy{We might need to define the problem or have a forward pointer.} \arash{See above}. Alexander et al. \cite{alexander2006integrality} showed the instance-specific integrality gap of the subtour elimination relaxation for the 2EC for instances of the problem with $n= 10$ is at most $\frac{7}{6}$.

@@

\section{Contributions of this paper}

>>>>

In combinatorial optimization the aim is to find the optimal solution in a discrete and
usually finite yet large set of solutions. For many specific combinatorial optimization problems such a solution can be found efficiently. For many others, finding optimal or in many cases near optimal solutions is NP-hard. A common approach to deal with such problems is relaxing the discrete solution set into a continuous set, where the optimization problem becomes tractable. Obtaining feasible solutions by means of such a relaxation requires an additional step of rounding the potentially fractional solution of the continuous relaxation into integer solutions.
}\fi

\section{Introduction}
In this paper we focus on finding feasible solutions to binary Integer Linear Programs (IP). Informally, an integer program is the optimization of a linear objective function subject to linear constraints, where the variables must take integer values. Binary variables represent yes/no decisions. Integer Programming (and more generally Mixed Integer Linear Programming (MILP)) can model many practical optimization problems including scheduling, logistics and resource allocation. It is  NP-hard even to determine if an IP instance has a feasible solution~\cite{GareyJohnson}.  However, there is substantial research into finding a feasible, provably-good approximate, and even (computationally) provably optimal solutions to specific IP instances. \cindy{TODO: probably need to informally distinguish between a IP for a general problem, and a specific instance when the parameters are set.}

A major tool for finding feasible solutions is the {\em linear-programming (LP) relaxation} for the instance. This is a new problem created by relaxing the integrality constraints for an IP instance, allowing the variables to take continuous (rational) values. Linear programs can be solved in polynomial time. The objective value of the linear programming relaxation provides a bound (lower bound for a minimization problem and upper bound for a maximization problem) on the optimal solution to the IP instance. The solutions can also provide some useful global structure, even though the fractional values might not be directly meaningful.

{\em LP-based approximation algorithms} use LP relaxations to find provably good approximate feasible solutions to IP problems in polynomial time. At the highest level, they involve solving the LP relaxation, using special structure from the problem to find a feasible solution, and proving that the objective value of the solution is no more than $C$ times worse than the bound from the LP relaxation. The approximation factor $C$ can be a constant or depend on the input parameters of the IP, e.g. $O(\log(n))$ where $n$ is the number of variables in the formulation of the IP (the dimension of the problem).

There is an inherent limit to how small $C$ can be for a given IP. The integrality gap for an IP instance is the ratio of the best integer solution to the best solution of the LP relaxation. Any LP-based approximation cannot have an approximation factor $C$ smaller than the integrality gap because there is no feasible solution with an objective value better than a factor of $C$ worse than the optimal solution of the LP relaxation.

If the integrality gap for an IP formulation is large, it is sometimes possible to add families of constraints to the formulation to reduce the integrality gap.  These constraints are redundant for the integer problem, but can make some fractional solutions no longer feasible for the LP. These families of constraints (cuts) can have exponential size as long as we can provide a polynomial-time separation algorithm. \cindy{To be self contained, we can add a few sentences about separation, but under the current time pressure, I won't for now.}

Reducing the integrality gap of an IP formulation has two advantages.  It can lead to better LP-based approximation algorithm bounds as described above.  It can also help exact solvers run faster or solve instances it could not before. Exact IP solvers are based on intelligent branch-and-bound strategies.  Commercial and open-source MILP solvers can find exact solutions (or near-optimal solutions with a provable bound) to many specific instances of NP-hard combinatorial optimization problems. These solvers use the LP relaxation to get lower bounds (for minimization problems).  The worst-case exponential search is practically feasible when the solver can prune large amounts of the search space.  This happens when the lower bound for a problem is worse than the value of a known feasible solution.  This requires a way to find good heuristic solution and it requires good lower bounds that are as close to the actual optimal value of an IP subproblem as possible.

In this paper, we give a method to find feasible solutions for IPs if the integrality gap is bounded. The method is also a tool for evaluating the integrality gap for a formulation.  Researcher can use it to determine whether they should expend effort to find new classes of cuts.  They can also use it to help guide theory for finding tighter bounds on the integrality gap for classic problems like the traveling salesman problem.

% A common approach to Integer programming is considering their linear relaxation.
For some problems such as the \textsc{Minimum Cost Spanning Tree Problem} there are linear programming relaxations whose basic feasible solutions coincide with  integral solutions, i.e. spanning trees.

%A common and successful approach is to round these (potentially) fractional solutions into integer solutions for the optimization problem at hand. The Integrality gap of a linear relaxation of an integer programming problem is the worst case ratio between the objective values of the discrete problem and the continuous problem. Equivalently, the integrality gap of the linear programming relaxation is a limit to the rounding approach: rounding a fractional solution into an integer solution incurs a multiplicative cost proportional to the integrality gap. In this paper we study integrality gaps for different combinatorial optimization problems and introduce new rounding algorithms that imply bounds on their respective integrality gaps.

@@

A polynomial time algorithm for proving an upper bound on integrality gap is called an LP-based approximation algorithm. For many well studied problems, we still do not know the exact integrality gap and the gap between the best known lower bound and the upper bound on the integrality gap are open. In some cases, there are known upper bounds, yet there is no known approximation algorithm, meaning that the proofs do not yield polynomial time algorithms.
}\fi 
The set of feasible points for a pure IP (henceforth IP) is the set
\begin{equation}
S(A,b)= \{x\in \mathbb{Z}^{n}\;:\; Ax\geq b\}  \label{S}.
\end{equation}
If we drop the integrality constraints, we have the linear relaxation of set $S(A,b)$,
\begin{equation}
P(A,b) = \{x\in \mathbb{R}^{n}\;:\; Ax\geq b\}. \label{P}
\end{equation}
Let $I=(A,b)$ denote an instance. Then $S(I)$ and $P(I)$ denote $S(A,b)$ and $P(A,b)$, respectively. Given a linear objective function $c$, an IP is $\min \;\{cx:\; x \in S(I)\}$.

Relaxing the integrality constraints gives the polynomial-time-solvable linear programming relaxation: $\min \;\{cx:\;x\in P(I) \}$.  The optimal value of this linear program (LP), denoted $z_{\lp}(I,c)$, is a lower bound on the optimal value for the IP, denoted $z_{\IP}(I,c)$.

Many researchers (see \cite{davids,vazirani}) have developed polynomial time LP-based approximation algorithms that find solutions for special classes of IPs whose cost are provably smaller than $C\cdot z_{LP}(I,c)$. However, for many combinatorial optimization problems there is a limit to such techniques based on LP relaxations, represented  by the {integrality gap} of the IP formulation. The integrality gap $g(I)$ for instance $I$ is defined to be $g(I)= \max_{c\geq 0}\frac{z_{IP}(I,c)}{z_{LP}(I,c)}$. \cindy{Does the integrality gap always require positive objective coefficients?  Or is that an extra restriction in our definition?  Is there a reference that has that restriction?}  For example consider the minimum cost 2-edge-connected multi-subgraph problem (2EC): Given a graph $G=(V,E)$ and $c\in \mathbb{R}^E_{\geq 0}$, 2EC asks for the minimum cost 2-edge-connected multi-subgraph of $G$. A linear programming relaxation for this problem known as the subtour elimination relaxation  is
 \begin{equation}\min \{cx: \sum_{e\in \delta(U)}x_e \geq 2 \mbox{ for } \emptyset \subsetneq U \subsetneq V,\; x\in [0,2]^{E}\}. 
 \end{equation}
 In this case the instance-specific integrality gap is the integrality gap of the subtour-elimination relaxation for the 2EC on graph with $n$ vertices. \cindy{We might need to define the problem or have a forward pointer.} \arash{See above}. Alexander et al. \cite{alexander2006integrality} showed the instance-specific integrality gap of the subtour elimination relaxation for the 2EC for instances of the problem with $n= 10$ is at most $\frac{7}{6}$.

@@

\subsection{Algorithms and Theory Contributions}
